Inferring the distribution and rate of Earth-like exoplanets \\ from noisy individual-planet inferences

### 2014-05-19 document/ms.tex
comments from @davidwhogg

This study is novel in a number of ways.
We create probabilistic information about the full exoplanet population
\emph{not} by weighting the detected objects by the inverse detection
efficiency (\citealt{petigura}), but rather by forward modeling the
observed distribution subject to the detection efficiency
(\dfmtodo{CITE TREMAINE?} \citealt{dong}).
Re-weighting the data by inverse selection probability---a method called
V-max in the quasar and galaxy luminosity function contexts---is not wrong,
but it produces a higher-variance estimate of the population than a justified
likelihood approach; it is higher variance because
the effective number of samples, after
weighting, can be much smaller than the actual number of samples.
In this work, we build a justified
likelihood of the observed data, taking into account the completeness function
or detection efficiency.
Our likelihood function is a variable-rate Poisson likelihood; it makes the
fundamental assumption that the data points are independently drawn from the
model, but is agnostic about all other properties of the data or distribution
functions.
Indeed, the assumptions behind this study here are weaker than those of any
previous study of the period and radius distribution of the Earth-sized
exoplanet population (\dfmtodo{CITE TREMAINE?} \citealt{dong, petigura}).

Another new aspect of this study is that it makes use of very flexible
distribution function models.
Indeed most of the models we consider are ``non-parametric'' in the weak sense
that they have enormous numbers of free parameters (not the strong sense of
having an \emph{infinite} number of free parameters).
The models are protected from degeneracies and over-fitting by priors.
In some models we assume that the two-dimensional distribution of planets in
period and radius is separable; in others we don't.
We treat all of these parameters as ``nuisance parameters'' when we ask
questions about the occurrence of Earth analogs.
That is, our rate conclusions are fully marginalized, and those
marginalizations are over non-trivial numbers of parameters.

When, in what follows, we ask questions about the radius disribution of
exoplanets, the distribution parameters are no longer nuisances The
radius-distribution parameters are what we are trying to measure.
Even in this case, however, our results will have the period-distribution
parameters marginalized out.

Perhaps the most important novelty of this study is in its treatment of
observational uncertainties.
Although planet periods in our data set are very precisely measured, the
planet radius measurements have large uncertainties.
The justified probabilistic approach to including these uncertainties in a
distribution analysis is to introduce the ``true'' radii as latent variables
and infer and marginalize them out along with all other nuisance parameters.
Although this sounds expensive---and it can be---there is a simple approach
we have been advocating (\cite{hogge}) that makes use of importance sampling.
We use this approach here; our results are fully marginalized over posterior
uncertainties in all of the planet radii.
More will be said about this below; the importance sampling looks like a
product over data points of sums (averages) over individual-datum radius
samplings.

@@

In this \paper, we are asking the question: \emph{what constraints can we put
on the occurence rate of exoplanets given all the light curves measured by
\kepler?}
Mathematically, this involves evaluating the so-called \emph{marginalized
likelihood function}
\begin{eqnarray}\eqlabel{crazylike}
p(\{\data_k\}\,|\,\ratepars) &=&
    \int p(\{\data_k\}\,|\,\{\entry_k\})
    \,p(\{\entry_k\}\,|\,\ratepars)
    \dd\{\entry_k\}
\end{eqnarray}
where $\{\data_k\}$ is the set of all light curves, one light curve $\data_k$
per target $k$, \ratepars\ is a vector of parameters describing the occurence
rate function $\rate_\ratepars(\entry)$ and $\entry_k$ is the vector of
physical parameters describing the planetary system (orbital periods,
radius ratios, stellar radius, \etc) around target $k$.
In this equation, our only assumption is that the datasets depend on the
occurence rate of exoplanets only through the catalog $\{\entry_k\}$.
In our case, this assumption qualitatively means that the signals found in the
light curves depend only on the actual planet properties and not on the
distributions from which they are drawn.
It is worth noting that---as we will discuss further below---the catalog only
provides \emph{probabilistic} constraints on $\{\entry_k\}$.

@@

In the inference procedure described above, we only discussed an abstract
occurence rate function.
To make it more concrete, we'll model the catalog as a Poisson draw from the
\emph{observable} rate function $\obs{\rate}_\ratepars$.
This leads to the well known result (see \citealt{tabachnik,youdin} for some
of the examples from the exoplanet literature)
\begin{eqnarray}\eqlabel{poisson-like}
p(\{\entry_k\}\,|\,\ratepars) &=&
    \exp\left(-\int \obs{\rate}_\ratepars (\entry) \dd\entry\right) \,
    \prod_{k=1}^K \obs{\rate}_\ratepars (\entry_k)\quad.
\end{eqnarray}
In this equation, the integral in the normalization term is the expected
number of observable exoplanets in the sample.
A further assumption that we have made in \eq{poisson-like} is that each
object is an independent sample from the rate function.
This assumption---though it is often made---cannot be correct when we consider
systems of multiple planets.
It is possible to relax this assumption (see \citealt{tremaine} for a related
technique) but this generalization is beyond the scope of the current \paper.

@@

As well as being a very flexible prior, this Gaussian process model is also
appealing because it is very well studied in the machine learning literature.
In particular, there is a very efficient algorithm---elliptical slice sampling
(ESS; \citealt{ess})---for generating samples from the posterior probability
function in \eq{posterior}.
As mentioned above, we actually go one step further and assume uniform
(hyper-)priors on $\ln\mean$ and $\ln\smoothpars$ in order to sample from the
joint posterior
\begin{eqnarray}
p(\ratepars,\,\mean,\,\smoothpars\,|\,\{\data_k\})
\end{eqnarray}
using Algorithm 1 from \citet{ess-hyper}.

@@

Given our domain knowledge of how detection efficiency depends on the physical
parameters, it might seem silly to measure the survey completeness in period
and radius instead of radius ratio or signal-to-noise.
It is also likely that a change of coordinates would yield a higher precision
result.
That being said, it is still correct to measure the completeness in period and
radius, and there are a few practical reasons for our choice.
The main argument is that since the radius uncertainties are dominated by
uncertainties in the stellar parameters, it is not possible to use the
published catalog (\citealt{petigura}) to compute constraints on radius
ratios.
In future catalogs, this problem would be solved by publishing representations
of all the posterior probabilities that are being measured.
In this case, the most useful data product would be \emph{posterior samples
for each target's radius ratio and stellar radius}.

@@

As well as inferring the ocurrence distribution of exoplanets, this dataset
can also be used to constrain the rate of Earth analogs.
Even though no true analog has been observed, we can make some assumptions
about the distribution (as described below) and extrapolate to the location of
the Earth.
Explicitly, we constrain the rate of exoplanets \emph{evaluated at the
location of Earth}
\begin{eqnarray}
\gammaearth &=& \rate (\ln\period_\oplus,\,\ln\radius_\oplus) \\
&=&
\left.\frac{\dd N}{\dd\ln\period\,\dd\ln\radius}\right|
_{\radius=\radius_\oplus,\,\period=\period_\oplus}\quad.
\end{eqnarray}
This is also the quantity that \citet{petigura} estimated in different units.

@@

We have made the first measurement of the Earth-like exoplanet population that
is based on forward modeling, takes proper account of the observational
uncertainties on planet radii, and makes use of extremely flexible
non-separable distributions for period and radius.
We find \hoggtodo{BLAH BLAH BLAH.}

Although our method makes weaker assumptions than any previous work in this
area, of course it \emph{does} make strong assumptions.
It assumes that the data set (from \citealt{petigura}) contains planets and
only planets (no false positives).
It assumes that the individual-planet radius uncertainties are also correctly
determined.
It assumes that the completeness calculation (again from \citealt{petigura})
is accurate and smoothly varying with period and radius; it makes no attempt
to adjust the completeness star-by-star for brightness or variability.
It assumes that the exoplanets are all independently drawn from the same
distribution function, with no regard to age, metallicity, or multiplicity.

Although we have taken account of the radius uncertainties in a principled
way, we did make one important approximation: We ignored planets with central
values far outside the period and radius box in which we built the model.
That is, we assumed that there are no very large outliers in radius
determination.
This assumption is a complex one; it involves the effective, end-to-end error
distribution in the data set, but also smoothness of the distribution function
at or near the edges of the analysis box.

If our inferences are right about true Earth analogs, then we expect that
there are about XXX transiting Earth analogs hidden in the \kepler\ data set.
We would be able to find some of these Earth analogs if the effective noise
level in the data could be reduced by a factor of about XXX.
We have identified a number of directions along which this noise reduction
might happen, including but not limited to retrospective data-driven or
physics-driven recalibrations of the \kepler\ focal plane
(\citealt{hoggwhitepaper}), new methods for performing aperture photometry,
and better models for stochastic instrument variability or stochastic
intrinsic stellar variability that take account of frequency structure
(\citealt{brewer, carter, roberts}).
If any (or all) of these deliver substantial improvements for \kepler\
archival data reanalyses, it's Nobel Prize time!  (Or at least \textsl{The
Colbert Report}?)

>>>>

@@

In this \paper, we are asking the question: \emph{what constraints can we put
on the occurence rate of exoplanets given all the light curves measured by
\kepler?}
Throughout this \paper, we use the notation $\rate_\ratepars(\entry)$ for the
occurence rate \rate---parameterized by the parameters \ratepars---as a
function of the physical parameters \entry\ (orbital period, planetary radius,
\etc).
In this framework, the occurence rate function can be ``parametric''---for
example, a power law---or a ``non-parametric'' function---such as a histogram
where the bin heights are the parameters \ratepars.
The light curve for a particular target $k$ is called $\data_k$ and the
\emph{true} set of physical parameters for that same object are given by
$\entry_k$.
A catalog provides a set of constraints on the parameters $\{\entry_k\}$
conditioned the set light curves $\{\data_k\}$.

Any inference about the rate function parameters can be formally addressed by
evaluating the \emph{marginalized likelihood function}
\begin{eqnarray}\eqlabel{crazylike}
p(\{\data_k\}\,|\,\ratepars) &=&
    \int p(\{\data_k\}\,|\,\{\entry_k\})
    \,p(\{\entry_k\}\,|\,\ratepars)
    \dd\{\entry_k\}
\end{eqnarray}
where, as we mentioned above, $\{\data_k\}$ is the set of all light curves,
one light curve $\data_k$ per target $k$, \ratepars\ is the vector of
parameters describing the occurence rate function $\rate_\ratepars(\entry)$
and $\entry_k$ is the vector of physical parameters describing the planetary
system (orbital periods, radius ratios, stellar radius, \etc) around target
$k$.
In this equation, our only assumption is that the datasets depend on the
occurence rate of exoplanets only through the catalog $\{\entry_k\}$.
In our case, this assumption qualitatively means that the signals found in the
light curves depend only on the actual planet properties and not on the
distributions from which they are drawn.
It is worth noting that---as we will discuss further below---the catalog only
provides \emph{probabilistic} constraints on $\{\entry_k\}$.

@@

In the inference procedure described above, we only discussed an abstract
occurence rate function.
To make it more concrete, we'll model the catalog as a Poisson draw from the
\emph{observable} rate function $\obs{\rate}_\ratepars$.
This leads to the previously known result (see \citealt{tabachnik,youdin} for
some of the examples from the exoplanet literature)
\begin{eqnarray}\eqlabel{poisson-like}
p(\{\entry_k\}\,|\,\ratepars) &=&
    \exp\left(-\int \obs{\rate}_\ratepars (\entry) \dd\entry\right) \,
    \prod_{k=1}^K \obs{\rate}_\ratepars (\entry_k)\quad.
\end{eqnarray}
In this equation, the integral in the normalization term is the expected
number of observable exoplanets in the sample.
A further assumption that we have made in \eq{poisson-like} is that each
object is an independent sample from the rate function.
This assumption---though it is often made---cannot be correct when we consider
systems of multiple planets.
It is possible to relax this assumption (see \citealt{tremaine} for a related
technique) but this generalization is beyond the scope of the current \paper.

@@

As well as being a very flexible prior, this Gaussian process model is also
appealing because it is well studied in the machine learning literature.
In particular, there is a very efficient algorithm---elliptical slice sampling
(ESS; \citealt{ess})---for generating samples from the posterior probability
function in \eq{posterior}.
As mentioned above, we actually go one step further and assume uniform
(hyper-)priors on $\ln\mean$ and $\ln\smoothpars$ in order to sample from the
joint posterior
\begin{eqnarray}
p(\ratepars,\,\mean,\,\smoothpars\,|\,\{\data_k\})
\end{eqnarray}
using Algorithm 1 from \citet{ess-hyper}.

@@

Given our domain knowledge of how detection efficiency depends on the physical
parameters, it might seem silly to measure the survey completeness in period
and radius instead of radius ratio or signal-to-noise.
It is also likely that a change of coordinates would yield a higher precision
result.
That being said, it is still correct to measure the completeness in period and
radius, and there are a few practical reasons for our choice.
The main argument is that since the radius uncertainties are dominated by
uncertainties in the stellar parameters, it is not possible to use the
published catalog (\citealt{petigura}) to compute constraints on radius
ratios.
In the future, this problem would be solved by publishing a representation of
\emph{the full posterior PDF for each object in the catalog}.
In this case, the most useful data product would be \emph{posterior samples
for each target's radius ratio and stellar radius}.

@@

As well as inferring the ocurrence distribution of exoplanets, this dataset
can also be used to constrain the rate of Earth analogs.
Even though no true analog has been observed, we can make some assumptions
about the distribution (as described below) and extrapolate to the location of
the Earth.
Explicitly, we constrain the rate of exoplanets \emph{evaluated at the
location of Earth}
\begin{eqnarray}
\gammaearth &=& \rate (\ln\period_\oplus,\,\ln\radius_\oplus) \\
&=&
\left.\frac{\dd N}{\dd\ln\period\,\dd\ln\radius}\right|
_{\radius=\radius_\oplus,\,\period=\period_\oplus}\quad.
\end{eqnarray}
This is also the quantity that \citet{petigura} estimated but in different
units.
That is, \gammaearth\ is the average number of exoplanets around a
``Sun-like'' star---per natural logarithmic bins in period and
radius---evaluated at the period and radius of Earth.

@@

\dfmtodo{write this section, eh?}

% We have made the first measurement of the Earth-like exoplanet population that
% is based on forward modeling, takes proper account of the observational
% uncertainties on planet radii, and makes use of extremely flexible
% non-separable distributions for period and radius.
% We find \hoggtodo{BLAH BLAH BLAH.}

% Although our method makes weaker assumptions than any previous work in this
% area, of course it \emph{does} make strong assumptions.
% It assumes that the data set (from \citealt{petigura}) contains planets and
% only planets (no false positives).
% It assumes that the individual-planet radius uncertainties are also correctly
% determined.
% It assumes that the completeness calculation (again from \citealt{petigura})
% is accurate and smoothly varying with period and radius; it makes no attempt
% to adjust the completeness star-by-star for brightness or variability.
% It assumes that the exoplanets are all independently drawn from the same
% distribution function, with no regard to age, metallicity, or multiplicity.

% Although we have taken account of the radius uncertainties in a principled
% way, we did make one important approximation: We ignored planets with central
% values far outside the period and radius box in which we built the model.
% That is, we assumed that there are no very large outliers in radius
% determination.
% This assumption is a complex one; it involves the effective, end-to-end error
% distribution in the data set, but also smoothness of the distribution function
% at or near the edges of the analysis box.

% If our inferences are right about true Earth analogs, then we expect that
% there are about XXX transiting Earth analogs hidden in the \kepler\ data set.
% We would be able to find some of these Earth analogs if the effective noise
% level in the data could be reduced by a factor of about XXX.
% We have identified a number of directions along which this noise reduction
% might happen, including but not limited to retrospective data-driven or
% physics-driven recalibrations of the \kepler\ focal plane
% (\citealt{hoggwhitepaper}), new methods for performing aperture photometry,
% and better models for stochastic instrument variability or stochastic
% intrinsic stellar variability that take account of frequency structure
% (\citealt{brewer, carter, roberts}).
% If any (or all) of these deliver substantial improvements for \kepler\
% archival data reanalyses, it's Nobel Prize time!  (Or at least \textsl{The
% Colbert Report}?)
